Pre-Analysis Plan: Ideological Biases of the Crowd?

### 2023-05-26 docs/pap.Rmd
add analyses and draft paper

---
title: "Pre-Analysis Plan: Ideological Biases of the Crowd?"
date: "`r format(Sys.time(), '%d %B, %Y - %X (%Z)')`"
output: 
  pdf_document:
    toc: yes
    number_sections: true
editor_options: 
  chunk_output_type: console
bibliography: references.bib
urlcolor: blue
linkcolor: blue
header-includes:
   - \usepackage{floatrow}
   - \floatsetup[figure]{capposition=top}
   - \floatsetup[table]{capposition=top}
   - \usepackage{booktabs}
   - \usepackage{amsmath}
   - \usepackage{longtable}
   - \usepackage{array}
   - \usepackage{multirow}
   - \usepackage{wrapfig}
   - \usepackage{float}
   - \usepackage{colortbl}
   - \usepackage{pdflscape}
   - \usepackage{tabu}
   - \usepackage{threeparttable}
   - \usepackage{threeparttablex}
   - \usepackage[normalem]{ulem}
   - \usepackage{makecell}
   - \usepackage{xcolor}
   - \usepackage[utf8]{inputenc}
---
```{r setup, include=FALSE}
## include this at top of your RMarkdown file for pretty output
## make sure to have the printr package installed: install.packages('printr')
knitr::opts_chunk$set(message = FALSE, warning = FALSE, echo = FALSE)
#library(printr)
library(tidyverse)
library(kableExtra)
options(kableExtra.latex.load_packages = FALSE)
```

@@

```{r conditions, topcaption=TRUE}
d <- tibble(Condition = c(rep("Specified", 4),
                          rep("Underspecified", 4)), 
            `Wording ENG` = c("[PVV/X] says immigration should be made harder.",
                              "[GreenLeft/X] says nitrogen emissions need to be reduced.",
                              "[Labour Party/X] says tax rate should go up for highest earners.",
                              "[Forum for Democracy/X] says that membership in the European Union has been especially bad for the Netherlands so far.".
                              "[PVV/X] says many immigrants are coming this way.",
                              "[GreenLeft/X] says nitrogen policy must be different.",
                              "[Labour Party/X] says tax system must be changed.",
                              "[Forum for Democracy/X] says the Netherlands should have a different role in the European Union."),
            `Wording NL` = c("[PVV/X] zegt dat immigratie moeilijker gemaakt moet worden.",
                             "[GroenLinks/X] zegt dat stikstofuitstoot meer tegengegaan moet worden.",
                             "[PvdA/X] zegt dat het belastingtarief voor de hoogste inkomens omhoog moet.",
                             "[Forum voor Democratie/X] zegt dat het lidmaatschap van de Europese Unie tot nu toe vooral slecht geweest voor Nederland is.",
                             "[PVV/X] zegt dat veel immigranten deze kant op komen.",
                             "[GroenLinks/X] zegt dat het stikstofbeleid anders moet.",
                             "[PvdA/X] zegt dat het belastingstelsel moet worden aangepast.",
                             "[Forum voor Democratie/X] zegt dat Nederland een andere rol in de Europese Unie moet hebben."))

@@

```{r demographics, topcaption=TRUE}
d <- tibble(Variable = c("gender (D1)", "age (D2)", 
                         "region (D3)", "job (D5)", "income (D6)",
                         "education (D7)"), 
            `Wording ENG` = c("What is your gender?", "What is your year of birth?",
                             "In which region do you live?",
                             "What applies most to you? Are you...",
                              "Can you indicate which income groups your net monthly household income falls into?",
                             "What is your highest level of education?", ),
            `Wording NL` = c("Wat is uw geslacht?", "Wat is uw geboortejaar?", 
                             "In welke regio woont u?",
                             "Wat is het meest op u van toepassing? Bent u...",
                             "Kunt u aangeven in welke inkomensgroepen uw netto maandelijks huishoudinkomen valt?", 
                             "Wat is uw hoogst GENOTEN opleiding",))

@@

```{r pret, topcaption=TRUE}
d <- tibble(Variable = c("vote recall (PT1)", "introduction",
                         "migration (PT2)", "climate (PT3)",
                         "tax (PT4)", "EU (PT5)", "RILE (PT6)",
                         "political-knowledge1 (PT7_1)", 
                         "political-knowledge2 (PT7_2)", 
                         "political-knowledge3 (PT7_3)", 
                         "political-knowledge4 (PT7_4)", 
                         "political-knowledge5 (PT7_5)",
                         "political-knowledge5 (PT7_6)"), 
            `Wording ENG` = c( "Which party did you vote for during the last parliamentary elections of March 2021?",
                               "To what extent do you agree with the following statements:",
                               "There are too many immigrants in the Netherlands.",
                               "The climate crisis is exaggerated.",
                               "The tax rate for the highest earners should go up",
                               ""
                               "In politics there is often talk about left and right. Where would you put yourself on this scale?",
                               "Which parties currently form the coalition government?",
                               "Which of the following is the current Finance Minister in the Rutte cabinet?",
                               "Which political party was the second-largest party in the last Lower House elections on March 17?",
                               "By whom are members of the Senate elected?",
                               "Who makes or is making new laws in the Netherlands?",
                               "If the government and the House of Representatives pass a law that is not in line with the Dutch Constitution, may the courts disapprove it?"),
            `Wording NL` = c("Op welke politieke partij heeft u bij de vorige Tweede Kamerverkiezingen van maart 2021 gestemd?",
                             "In hoeverre bent u het eens met onderstaande stellingen:",
                             "Er zijn te veel immigranten in Nederland.",
                             "De klimaatcrisis wordt overdreven.",
                             "Het belastingtarief voor de hoogste inkomens moet omhoog",
                             "Het lidmaatschap van de Europese Unie is tot nu toe vooral slecht geweest voor Nederland",
                             "In de politiek wordt vaak gesproken over links en rechts. Waar zou u uzelf op deze schaal plaatsen?",
                             "Welke partijen vormen op dit moment de coalitieregering?",
                            "Wie van de volgende personen is de huidige minister van Financiën in het kabinet Rutte?",
                            "Welke politieke partij was de op één na grootste partij tijdens de afgelopen Tweede Kamerverkiezingen van 17 maart?",
                            "Door wie worden de leden van de Eerste Kamer gekozen?",
                            "Wie maakt of maken in Nederland nieuwe wetten?",
                            "Als de regering en de Tweede Kamer een wet aannemen die niet overeenkomt met de Nederlandse Grondwet, mag de rechter deze wet dan afkeuren?"))

@@

## Balance Checks
We will conduct a balance test based on demographics (age, gender, education, geographical region, level of urbanness,employment, and income), vote choice in the 2021 parliamentary elections, ideological self-placement, political knowledge, and positions on the issues, using the `cobalt` R package [@greifer2021]. 
If the groups are unbalanced on one of these variables -- i.e. standardized mean differences $<0.05$ -- we will add the covariates to the analyses.
We will use the code below to conduct the balance tests (see [here](https://github.com/MarikenvdVelden/compromise-punish/tree/main/src/analysis/balance-test.R) for the R script).

@@

## Hypothesis 1
To test the _Ideological Bias hypothesis_, we will conduct a multilevel model, with respondents clustered in issues, see code chunck below.
Using the pooled data we will estimate a within groups fixed effects model. 
$Y\hat{Y}_{r, i, t}$ in Equation \ref{eq:pooled} denotes the  evaluation  of  a stance by respondent $r$, during issue $i$ and at experimental round $t$ -- ranging from round 1 to round 4.
The standard errors are clustered at the individual level.\

$$\begin{equation}\label{eq:pooled}
    \hat{stance-correct}_{r, i, t} =\beta_{0} + \beta_{1}masked_{r, i, t} +
    \beta_{2}specification_{r, i, t} + 
    \beta_{3}ideological-distance-to-party_{r, i, t} +
    \beta_{4}political-knowledge_{r, i, t} +
    \alpha_{i} + \gamma_{t} + \varepsilon_{r, i, t}
\end{equation}$$

@@

## Hypothesis 2
To test the _Overinterpretation Hypotheses_, we will conduct a multilevel model, with respondents clustered in issues, see code chunck below.
Using the pooled data we will estimate a within groups fixed effects model. 
$Y\hat{Y}_{r, i, t}$ in Equation \ref{eq:pooled} denotes the  evaluation  of  a stance by respondent $r$, during issue $i$ and at experimental round $t$ -- ranging from round 1 to round 4.
The standard errors are clustered at the individual level

$$\begin{equation}\label{eq:pooled}
    \hat{stance-correct}_{r, i, t} =\beta_{0} + \beta_{1}masked_{r, i, t} +
    \beta_{2}specification_{r, i, t} + 
    \beta_{3}heterogeneous_treatment{r, i, t} +
    \beta_{4} specification * heterogeneous_treatment_{r, i, t} +
    \alpha_{i} + \gamma_{t} + \varepsilon_{r, i, t}
\end{equation}$$

@@

## Hypothesis 3
To test the _Masking Solution hypotheses_, we will conduct a multilevel model, with respondents clustered in issues, see code chunck below.
Using the pooled data we will estimate a within groups fixed effects model. 
$Y\hat{Y}_{r, i, t}$ in Equation \ref{eq:pooled} denotes the  evaluation  of  a stance by respondent $r$, during issue $i$ and at experimental round $t$ -- ranging from round 1 to round 4.
The standard errors are clustered at the individual level

$$\begin{equation}\label{eq:pooled}
    \hat{stance-correct}_{r, i, t} =\beta_{0} + \beta_{1}masked_{r, i, t} +
    \beta_{2}specification_{r, i, t} + 
    \beta_{3}heterogeneous_treatment{r, i, t} +
    \beta_{4} masked * heterogeneous_treatment_{r, i, t} +
    \alpha_{i} + \gamma_{t} + \varepsilon_{r, i, t}
\end{equation}$$

@@

## Statistical Significance
All the hypotheses are directional, and therefore all of the tests will be one-tailed.
We will use an $\alpha$-value of $0.05$ as the value for statistical significance in all models above.

>>>>

---
title: "Pre-Analysis Plan: Ideological Biases of the Crowd?"
date: "`r format(Sys.time(), '%d %B, %Y - %X (%Z)')`"
output: 
  pdf_document:
    toc: yes
    number_sections: true
editor_options: 
  chunk_output_type: console
bibliography: references.bib
urlcolor: blue
linkcolor: blue
header-includes:
   - \usepackage{floatrow}
   - \floatsetup[figure]{capposition=top}
   - \floatsetup[table]{capposition=top}
   - \usepackage{booktabs}
   - \usepackage{amsmath}
   - \usepackage{longtable}
   - \usepackage{array}
   - \usepackage{multirow}
   - \usepackage{wrapfig}
   - \usepackage{float}
   - \usepackage{colortbl}
   - \usepackage{pdflscape}
   - \usepackage{tabu}
   - \usepackage{threeparttable}
   - \usepackage{threeparttablex}
   - \usepackage[normalem]{ulem}
   - \usepackage{makecell}
   - \usepackage{xcolor}
   - \usepackage[utf8]{inputenc}
---
```{r setup, include=FALSE}
## include this at top of your RMarkdown file for pretty output
## make sure to have the printr package installed: install.packages('printr')
knitr::opts_chunk$set(message = FALSE, warning = FALSE, echo = FALSE,cache = TRUE)
#library(printr)
library(tidyverse)
library(kableExtra)
options(kableExtra.latex.load_packages = FALSE)
```

@@

```{r conditions, topcaption=TRUE}
d <- tibble(Condition = c(rep("Specified", 4),
                          rep("Underspecified", 4)), 
            `Wording ENG` = c("[PVV/X] says immigration should be made harder.",
                              "[GreenLeft/X] says nitrogen emissions need to be reduced.",
                              "[Labour Party/X] says tax rate should go up for highest earners.",
                              "[Forum for Democracy/X] says that membership in the European Union has been especially bad for the Netherlands so far.",
                              "[PVV/X] says many immigrants are coming this way.",
                              "[GreenLeft/X] says nitrogen policy must be different.",
                              "[Labour Party/X] says tax system must be changed.",
                              "[Forum for Democracy/X] says the Netherlands should have a different role in the European Union."),
            `Wording NL` = c("[PVV/X] zegt dat immigratie moeilijker gemaakt moet worden.",
                             "[GroenLinks/X] zegt dat stikstofuitstoot meer tegengegaan moet worden.",
                             "[PvdA/X] zegt dat het belastingtarief voor de hoogste inkomens omhoog moet.",
                             "[Forum voor Democratie/X] zegt dat het lidmaatschap van de Europese Unie tot nu toe vooral slecht geweest voor Nederland is.",
                             "[PVV/X] zegt dat veel immigranten deze kant op komen.",
                             "[GroenLinks/X] zegt dat het stikstofbeleid anders moet.",
                             "[PvdA/X] zegt dat het belastingstelsel moet worden aangepast.",
                             "[Forum voor Democratie/X] zegt dat Nederland een andere rol in de Europese Unie moet hebben."))

@@

```{r demographics, topcaption=TRUE}
d <- tibble(Variable = c("gender (D1)", "age (D2)", 
                         "region (D3)", "job (D5)", "income (D6)",
                         "education (D7)"), 
            `Wording ENG` = c("What is your gender?", 
                              "What is your year of birth?",
                             "In which region do you live?",
                             "What applies most to you? Are you...",
                              "Can you indicate which income groups your net monthly household income falls into?",
                             "What is your highest level of education?"),
            `Wording NL` = c("Wat is uw geslacht?", "Wat is uw geboortejaar?", 
                             "In welke regio woont u?",
                             "Wat is het meest op u van toepassing? Bent u...",
                             "Kunt u aangeven in welke inkomensgroepen uw netto maandelijks huishoudinkomen valt?", 
                             "Wat is uw hoogst GENOTEN opleiding"))

@@

```{r pret, topcaption=TRUE}
d <- tibble(Variable = c("vote recall (PT1)", "introduction",
                         "migration (PT2)", "climate (PT3)",
                         "tax (PT4)", "EU (PT5)", "RILE (PT6)",
                         "political-knowledge1 (PT7_1)", 
                         "political-knowledge2 (PT7_2)", 
                         "political-knowledge3 (PT7_3)", 
                         "political-knowledge4 (PT7_4)", 
                         "political-knowledge5 (PT7_5)",
                         "political-knowledge5 (PT7_6)"), 
            `Wording ENG` = c( "Which party did you vote for during the last parliamentary elections of March 2021?",
                               "To what extent do you agree with the following statements:",
                               "There are too many immigrants in the Netherlands.",
                               "The climate crisis is exaggerated.",
                               "The tax rate for the highest earners should go up",
                               "The Netherlands should have a different role in the European Union.",
                               "In politics there is often talk about left and right. Where would you put yourself on this scale?",
                               "Which parties currently form the coalition government?",
                               "Which of the following is the current Finance Minister in the Rutte cabinet?",
                               "Which political party was the second-largest party in the last Lower House elections on March 17?",
                               "By whom are members of the Senate elected?",
                               "Who makes or is making new laws in the Netherlands?",
                               "If the government and the House of Representatives pass a law that is not in line with the Dutch Constitution, may the courts disapprove it?"),
            `Wording NL` = c("Op welke politieke partij heeft u bij de vorige Tweede Kamerverkiezingen van maart 2021 gestemd?",
                             "In hoeverre bent u het eens met onderstaande stellingen:",
                             "Er zijn te veel immigranten in Nederland.",
                             "De klimaatcrisis wordt overdreven.",
                             "Het belastingtarief voor de hoogste inkomens moet omhoog",
                             "Het lidmaatschap van de Europese Unie is tot nu toe vooral slecht geweest voor Nederland",
                             "In de politiek wordt vaak gesproken over links en rechts. Waar zou u uzelf op deze schaal plaatsen?",
                             "Welke partijen vormen op dit moment de coalitieregering?",
                            "Wie van de volgende personen is de huidige minister van Financiën in het kabinet Rutte?",
                            "Welke politieke partij was de op één na grootste partij tijdens de afgelopen Tweede Kamerverkiezingen van 17 maart?",
                            "Door wie worden de leden van de Eerste Kamer gekozen?",
                            "Wie maakt of maken in Nederland nieuwe wetten?",
                            "Als de regering en de Tweede Kamer een wet aannemen die niet overeenkomt met de Nederlandse Grondwet, mag de rechter deze wet dan afkeuren?"))

@@

## Balance Checks
We will conduct a balance test based on demographics (age, gender, education, geographical region, level of urbanness,employment, and income), vote choice in the 2021 parliamentary elections, ideological self-placement, political knowledge, and positions on the issues, using the `cobalt` R package [@greifer2021]. 
If the groups are unbalanced on one of these variables -- i.e. standardized mean differences $<0.05$ -- we will add the covariates to the analyses.
We will use the code below to conduct the balance tests (see [here](https://github.com/MarikenvdVelden/bias-experiment/blob/master/src/analysis/balance-test.R) for the R script).

@@

## Hypothesis 1
To test the _Ideological Bias hypothesis_, we will conduct a multilevel model, with respondents clustered in issues, see code chunck below.
Using the pooled data we will estimate a within groups fixed effects model. 
$Y\hat{Y}_{r, i, t}$ in Equation \ref{eq:pooled1} denotes the  evaluation  of  a stance by respondent $r$, during issue $i$ and at experimental round $t$ -- ranging from round 1 to round 4.
The standard errors are clustered at the individual level.\

\begin{equation}\label{eq:pooled1}
    \hat{stance-correct}_{r, i, t} =\beta_{0} + \beta_{1}masked_{r, i, t} +
    \beta_{2}specification_{r, i, t} + 
    \beta_{3}ideological-distance-to-party_{r, i, t} +
    \beta_{4}political-knowledge_{r, i, t} +
    \alpha_{i} + \gamma_{t} + \varepsilon_{r, i, t}
\end{equation}

@@

## Hypothesis 2
To test the _Overinterpretation Hypotheses_, we will conduct a multilevel model, with respondents clustered in issues, see code chunck below.
Using the pooled data we will estimate a within groups fixed effects model. 
$Y\hat{Y}_{r, i, t}$ in Equation \ref{eq:pooled2} denotes the  evaluation  of  a stance by respondent $r$, during issue $i$ and at experimental round $t$ -- ranging from round 1 to round 4.
The standard errors are clustered at the individual level

\begin{equation}\label{eq:pooled2}
    \hat{stance-correct}_{r, i, t} =\beta_{0} + \beta_{1}masked_{r, i, t} +
    \beta_{2}specification_{r, i, t} + 
    \beta_{3}heterogeneous_treatment{r, i, t} +
    \beta_{4} specification * heterogeneous_treatment_{r, i, t} +
    \alpha_{i} + \gamma_{t} + \varepsilon_{r, i, t}
\end{equation}

@@

## Hypothesis 3
To test the _Masking Solution hypotheses_, we will conduct a multilevel model, with respondents clustered in issues, see code chunck below.
Using the pooled data we will estimate a within groups fixed effects model. 
$Y\hat{Y}_{r, i, t}$ in Equation \ref{eq:pooled3} denotes the  evaluation  of  a stance by respondent $r$, during issue $i$ and at experimental round $t$ -- ranging from round 1 to round 4.
The standard errors are clustered at the individual level

\begin{equation}\label{eq:pooled3}
    \hat{stance-correct}_{r, i, t} =\beta_{0} + \beta_{1}masked_{r, i, t} +
    \beta_{2}specification_{r, i, t} + 
    \beta_{3}heterogeneous_treatment{r, i, t} +
    \beta_{4} masked * heterogeneous_treatment_{r, i, t} +
    \alpha_{i} + \gamma_{t} + \varepsilon_{r, i, t}
\end{equation}

@@

## Statistical Significance
All the hypotheses are directional, and therefore all of the tests will be one-tailed.
We will use an $\alpha$-value of ``0.05` as the value for statistical significance in all models above.
